---
title: "Analysis: CrispRVariants"
subtitle: "Brzezinski - Purvis Dev paper CRISPR editing outcomes"
author: 
  name: "Michael Kaufman, PhD"
  orcid: 0000-0003-2441-5836
  email: michael.kaufman@cuanschutz.edu
  affiliation: 
    name: "University of Colorado Cancer Center"
    url: "https://medschool.cuanschutz.edu/colorado-cancer-center"
abstract:
  ""

editor: source
theme: cosmo
number-sections: true
link-citations: true
link-external-references: true

format:
  html:
    title-block-banner: "report_assets/analysis-logo.png"
    embed-resources: true
    self_contained: true
    highlight-style: pygments
    code-fold: true
    code-summary: "[code]"
    code-overflow: wrap
    toc: true
    toc-depth: 2
    toc-location: right  
    css: report_assets/custom-toc.css
    grid:
      sidebar-width: 300px
      body-width: 1600px
      margin-width: 300px
      gutter-width: 10px

execute:
  warning: false
  message: false
  fig.align: "center"

params:
  bwa_threads: 6
  outs_path: "outs/01-CrispRVariants"
  fastq_path: "../raw_data/fastq"
  bam_path: "../processed_data"
  ref_genome: "../ref/mm39.fa"
  gtf_file: "../ref/gencode.vM37.basic.annotation.gtf"
  treatment: "TestDHS15primers_R2"
  control: "ControlDHS15primers_R2"
  guide_name: "DHS15g2" # DHS15g1 or DHS15g2
  guide_sequence: "TTTTCTTTTTTATTTAACCG" # GCTGCCCCAGCCTTTCACAA or TTTTCTTTTTTATTTAACCG
  target_chr: "chr14"
  target_padding: 30
---

## Notes

This document is part of the Purvis Dev 2025 CRISPR editing outcomes analysis. It uses the `CrispRVariants` package to analyze CRISPR editing outcomes from paired-end sequencing data. Amplicon sequencing data is used to assess the efficiency of CRISPR editing at a specific target region in the mouse genome.

The following external tools are used in this analysis and must be installed and available in your PATH:
`bwa mem` is used for alignment, and `CrispRVariants` is used to analyze the editing outcomes.
`samtools` is used for BAM file processing.

This experiment uses paired-end sequencing data, where each sample has two FASTQ files (R1 and R2). The analysis will align these reads to a reference genome, create a `CrispRVariants` object, and visualize the editing outcomes.

## Environment

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align='center')

# Start timing the document render
start_time <- Sys.time()
```

### Load libraries

```{r load libraries}

library(CrispRVariants)
library(Rsamtools)
library(GenomicRanges)
library(GenomicFeatures)
library(Biostrings)
library(DT)
library(gridExtra)
library(Gviz)

source("scripts/qmd_embed_download.R")
source("scripts/renv_embed_download.R")

set.seed(87645893)

dir.create(params$outs_path, showWarnings = TRUE)
```

## Analysis parameters

```{r samples}
#datatable(t(as.data.frame(params)), 
#          options = list(paging = FALSE))
```

## Alignment of FASTQ files

This step calls the BWA and SAMtools commands to align the FASTQ files to the reference genome and create sorted BAM files.

```{r align-fastq}
dir.create(params$bam_path, recursive = TRUE, showWarnings = FALSE)

fastqs <- list.files(params$fastq_path, pattern="\\.(fastq|fq)(\\.gz)?$", full.names=TRUE, ignore.case=TRUE)
stopifnot(length(fastqs) > 0)

for (fq in fastqs) {
  sample <- sub("\\.(fastq|fq)(\\.gz)?$", "", basename(fq), ignore.case=TRUE)
  bam <- file.path(params$bam_path, paste0(sample, "_sorted.bam"))

  if (!file.exists(bam)) {
    message("Aligning: ", sample)
    cmd <- sprintf('bwa mem -t %d "%s" "%s" | samtools sort -@ %d -o "%s"',
                   as.integer(params$bwa_threads), params$ref_genome, fq,
                   as.integer(params$bwa_threads), bam)
    status <- system(cmd)
    if (status != 0) stop("Alignment failed for: ", sample)
    system2("samtools", c("index", "-@", as.character(params$bwa_threads), bam))
  } else {
    message("Skipping ", sample, " — BAM already exists.")
  }
}
```

## CrispR editing outcomes analysis

### Create the target location and reference sequence

```{r auto-detect-target}
# Auto-detect target region based on guide sequence
detect_target_from_guide <- function(ref_genome, guide_seq, target_chr = NULL, flank_size = 50) {
  
  message("Searching for guide sequence: ", guide_seq)
  
  chromosomes_to_search <- if(!is.null(target_chr)) {
    target_chr
  } else {
    chr_cmd <- sprintf("samtools faidx %s.fai | cut -f1", ref_genome)
    chrs <- system(chr_cmd, intern = TRUE)
    chrs[grepl("^chr", chrs)]
  }
  
  target_locations <- list()
  
  for(chr in chromosomes_to_search) {
    message("Searching chromosome: ", chr)
    
    # Extract chromosome sequence
    chr_seq_cmd <- sprintf("samtools faidx %s %s", ref_genome, chr)
    chr_lines <- system(chr_seq_cmd, intern = TRUE)
    
    if(length(chr_lines) < 2) next
    
    chr_seq <- paste(chr_lines[-1], collapse = "")
    chr_dna <- DNAString(chr_seq)
    
    # Search for guide sequence on forward strand
    forward_matches <- matchPattern(guide_seq, chr_dna)
    if(length(forward_matches) > 0) {
      for(i in seq_along(forward_matches)) {
        match_start <- start(forward_matches)[i]
        match_end <- end(forward_matches)[i]
        
        # For forward strand: cut site is typically 3bp upstream of PAM
        # Assuming PAM is immediately after guide (position 21-23 of guide+PAM)
        cut_site <- match_end - 3  # 3bp upstream of PAM (which would be after guide)
        
        target_locations[[length(target_locations) + 1]] <- list(
          chr = chr,
          guide_start = match_start,
          guide_end = match_end,
          cut_site = cut_site,
          strand = "+",
          target_start = max(1, cut_site - flank_size),
          target_end = cut_site + flank_size,
          target_loc_forward = cut_site - max(1, cut_site - flank_size) + 1,
          target_loc_reverse = cut_site - max(1, cut_site - flank_size) + 1
        )
      }
    }
    
    # Search for guide sequence on reverse strand
    reverse_matches <- matchPattern(reverseComplement(DNAString(guide_seq)), chr_dna)
    if(length(reverse_matches) > 0) {
      for(i in seq_along(reverse_matches)) {
        match_start <- start(reverse_matches)[i]
        match_end <- end(reverse_matches)[i]
        
        # For reverse strand: cut site is 3bp downstream of PAM
        # The guide we found is the reverse complement, so cut site is near the start
        cut_site <- match_start + 3
        
        target_locations[[length(target_locations) + 1]] <- list(
          chr = chr,
          guide_start = match_start,
          guide_end = match_end,
          cut_site = cut_site,
          strand = "-",
          target_start = max(1, cut_site - flank_size),
          target_end = cut_site + flank_size,
          target_loc_forward = cut_site - max(1, cut_site - flank_size) + 1,
          target_loc_reverse = cut_site - max(1, cut_site - flank_size) + 1
        )
      }
    }
  }
  
  return(target_locations)
}

# Auto-detect target region
detected_targets <- detect_target_from_guide(
  ref_genome = params$ref_genome,
  guide_seq = params$guide_sequence,
  target_chr = params$target_chr,  # Use specified chromosome if available
  flank_size = params$target_padding  # 30bp flanking regions
)

message("Found ", length(detected_targets), " potential target sites")

if(length(detected_targets) == 0) {
  stop("Guide sequence not found in reference genome!")
}

# Display detected targets
detected_df <- do.call(rbind, lapply(detected_targets, function(x) {
  data.frame(
    Chromosome = x$chr,
    Guide_Start = x$guide_start,
    Guide_End = x$guide_end,
    Cut_Site = x$cut_site,
    Strand = x$strand,
    Target_Start = x$target_start,
    Target_End = x$target_end,
    stringsAsFactors = FALSE
  )
}))

DT::datatable(detected_df, caption = "Detected Target Locations for Guide Sequence")

# Use the first detected target (or filter by chromosome if specified)
if(!is.null(params$target_chr)) {
  # Filter to specified chromosome
  chr_matches <- detected_targets[sapply(detected_targets, function(x) x$chr == params$target_chr)]
  if(length(chr_matches) > 0) {
    selected_target <- chr_matches[[1]]
  } else {
    stop("No targets found on specified chromosome: ", params$target_chr)
  }
} else {
  selected_target <- detected_targets[[1]]
}

message("Selected target:")
message("  Chromosome: ", selected_target$chr)
message("  Target region: ", selected_target$target_start, "-", selected_target$target_end)
message("  Strand: ", selected_target$strand)
message("  Cut site: ", selected_target$cut_site)

# Override parameters with detected values
auto_params <- list(
  target_chr = selected_target$chr,
  target_start = selected_target$target_start,
  target_end = selected_target$target_end,
  target_strand = selected_target$strand,
  target_loc_forward = selected_target$target_loc_forward,
  target_loc_reverse = selected_target$target_loc_reverse
)
```

```{r target-location}
# Define the target location and reference sequence as granges
target_location <- GRanges(
  seqnames = auto_params$target_chr,
  ranges = IRanges(start = auto_params$target_start, end = auto_params$target_end),
  strand = auto_params$target_strand
)

# Extract reference sequence
ref_cmd <- sprintf("samtools faidx %s %s:%s-%s",
                  params$ref_genome,
                  seqnames(target_location)[1],
                  start(target_location)[1],
                  end(target_location)[1])

ref_lines <- system(ref_cmd, intern = TRUE)

if(length(ref_lines) < 2) {
  stop("Failed to extract reference sequence")
}

reference_raw <- paste(ref_lines[-1], collapse = "")

# Use the detected strand to orient the reference sequence correctly
if(auto_params$target_strand == "-") {
  reference <- Biostrings::reverseComplement(Biostrings::DNAString(reference_raw))
  message("Using reverse complement for minus strand target")
  message("Original sequence: ", substr(reference_raw, 1, 50), "...")
  message("Rev comp sequence: ", substr(as.character(reference), 1, 50), "...")
} else {
  reference <- Biostrings::DNAString(reference_raw)
  message("Using forward strand reference sequence")
}

# Validate guide sequence is found in the oriented reference
guide_found_in_reference <- grepl(params$guide_sequence, as.character(reference), fixed = TRUE)

if(guide_found_in_reference) {
  message("✓ Guide sequence found in oriented reference sequence")
  
  # Find the position of the guide in the reference for validation
  guide_pos <- regexpr(params$guide_sequence, as.character(reference), fixed = TRUE)
  if(guide_pos[1] != -1) {
    message("Guide found at position ", guide_pos[1], " in reference sequence")
  }
} else {
  warning("✗ Guide sequence NOT found in oriented reference sequence!")
  message("This suggests the reference orientation may be incorrect")
  message("Guide sequence: ", params$guide_sequence)
  message("Reference sequence: ", substr(as.character(reference), 1, 100), "...")
}

message("Target strand: ", auto_params$target_strand)
message("Target width: ", width(target_location))
message("Reference length: ", nchar(as.character(reference)))
message("Guide sequence: ", params$guide_sequence)

# Show final parameters being used
final_params_df <- data.frame(
  Parameter = c("target_chr", "target_start", "target_end", "target_strand", 
               "target_loc_forward", "target_loc_reverse"),
  Value = c(auto_params$target_chr, auto_params$target_start, auto_params$target_end, 
           auto_params$target_strand, auto_params$target_loc_forward, auto_params$target_loc_reverse),
  stringsAsFactors = FALSE
)

DT::datatable(final_params_df, caption = "Final Target Parameters")
```

### CrisprSet object

This object is created from the sorted BAM files and the target location. It will contain the read counts and variant information for each sample at the specified target region.

```{r crispr-set}
# Get all sorted BAM files first
all_bam_files <- list.files(params$bam_path, pattern = "_sorted\\.bam$", full.names = TRUE)

if(length(all_bam_files) == 0) {
  stop("No sorted BAM files found in ", params$bam_path, ". Check alignment step.")
}

all_sample_names <- gsub("_sorted\\.bam$", "", basename(all_bam_files))

# Filter to only treatment and control samples based on parameters
treatment_indices <- which(grepl(params$treatment, all_sample_names))
control_indices <- which(grepl(params$control, all_sample_names))

# Combine indices and get corresponding BAM files
selected_indices <- c(treatment_indices, control_indices)
selected_bam_files <- all_bam_files[selected_indices]
selected_sample_names <- all_sample_names[selected_indices]

message("All BAM files found: ", length(all_bam_files))
message("Treatment pattern '", params$treatment, "' matches: ", length(treatment_indices), " files")
message("Control pattern '", params$control, "' matches: ", length(control_indices), " files")
message("Selected BAM files for analysis: ", length(selected_bam_files))

if(length(selected_bam_files) == 0) {
  stop("No BAM files match the treatment (", params$treatment, ") or control (", params$control, ") patterns!")
}

message("Selected BAM files:")
for(i in seq_along(selected_bam_files)) {
  sample_type <- if(i <= length(treatment_indices)) "TREATMENT" else "CONTROL"
  message("  - ", basename(selected_bam_files[i]), " (", sample_type, ")")
}

# Check selected files exist and are indexed
for(bam_file in selected_bam_files) {
  if(!file.exists(bam_file)) {
    stop("BAM file not found: ", bam_file)
  }
  if(!file.exists(paste0(bam_file, ".bai"))) {
    message("Indexing missing for: ", basename(bam_file))
    indexBam(bam_file)
  }
}

# Use ONLY the selected BAM files for CrisprSet creation
crispr_set <- readsToTarget(selected_bam_files, 
                           target = target_location, 
                           reference = reference,
                           names = selected_sample_names, 
                           target.loc = auto_params$target_loc_forward)

DT::datatable(variantCounts(crispr_set))

# Update the global variables to use selected samples
sorted_bam_files <- selected_bam_files
bam_sample_names <- selected_sample_names
```

### Variant analysis

This section creates variant plots for the CRISPR editing outcomes. It uses the `CrispRVariants` package to visualize the variants detected in the target region across samples.

```{r variant-plots}
gtf_fname <- params$gtf_file

# Check if GTF file exists
if(file.exists(gtf_fname)) {
  message("Creating TxDb from GTF file...")
  txdb <- GenomicFeatures::makeTxDbFromGFF(gtf_fname, format = "gtf")
  
  # Check if there are any genes in our target region
  target_genes <- genes(txdb, filter = list(tx_chrom = seqnames(target_location)[1]))
  overlapping_genes <- subsetByOverlaps(target_genes, target_location)
  
  if(length(overlapping_genes) == 0) {
    message("No genes found overlapping target region. Proceeding without gene annotation.")
    txdb <- NULL
  } else {
    message("Found ", length(overlapping_genes), " genes overlapping target region")
    print(overlapping_genes)
  }
} else {
  message("GTF file not found at: ", gtf_fname)
  message("Proceeding without gene annotation")
  txdb <- NULL
}
```


```{r variant-plots2}
#| fig.width: 16
#| fig.height: 16
#| results: "hide"
#| message: false
#| warning: false

# Now group will only include the selected samples
group <- ifelse(grepl(params$treatment, bam_sample_names), "treatment", 
                ifelse(grepl(params$control, bam_sample_names), "control", "other"))

# Validation - should now only show selected samples
message("Sample groupings (selected samples only):")
message("Treatment (", params$treatment, "): ", paste(bam_sample_names[group == "treatment"], collapse = ", "))
message("Control (", params$control, "): ", paste(bam_sample_names[group == "control"], collapse = ", "))
message("Other: ", paste(bam_sample_names[group == "other"], collapse = ", "))

# Verify we have both groups
if(!any(group == "treatment")) {
  warning("No treatment samples found in selected BAMs!")
}
if(!any(group == "control")) {
  warning("No control samples found in selected BAMs!")
}

# Use the main parameter in plotVariants instead of trying to add title with +
plotVariants(crispr_set, txdb = txdb, gene.text.size = 8,
    main = paste0("Variant Counts for ", params$guide_name, ": ", params$guide_sequence),
    row.ht.ratio = c(1,8), col.wdth.ratio = c(4,2),
    plotAlignments.args = list(line.weight = 0.5, ins.size = 2, 
                               legend.symbol.size = 4),
    plotFreqHeatmap.args = list(plot.text.size = 3, x.size = 8, group = group, 
                                legend.text.size = 8, 
                                legend.key.height = grid::unit(0.5, "lines")))

# Use parameter-based filename
pdf(file.path(params$outs_path, paste0(params$treatment, "_vs_", params$control, "_variant_plot.pdf")), 
    width = 16, height = 16)
plotVariants(crispr_set, 
            main = paste0("Variant Counts for ", params$guide_name, ": ", params$guide_sequence),
            txdb = txdb, gene.text.size = 8)
dev.off()
```

```{r}
#| fig.width: 8
#| fig.height: 8
plotFreqHeatmap(crispr_set, type = "proportions") + labs(title = paste0("Proportions of Variant Frequency for ", params$guide_name, ":", params$guide_sequence))
```

```{r mutation-efficiency}
# Calculate mutation efficiency
control_pattern <- params$control
treatment_pattern <- params$treatment

# Dynamic sample identification using the parameters
treatment_samples <- bam_sample_names[grepl(treatment_pattern, bam_sample_names)]
control_samples <- bam_sample_names[grepl(control_pattern, bam_sample_names)]

message("Treatment samples: ", paste(treatment_samples, collapse = ", "))
message("Control samples: ", paste(control_samples, collapse = ", "))

if(length(treatment_samples) > 0 && length(control_samples) > 0) {
  tryCatch({
    eff <- mutationEfficiency(crispr_set, 
                             filter.cols = control_pattern, 
                             exclude.cols = control_pattern)
    message("Mutation efficiency: ", round(eff, 4))
    
    # Create summary table
    efficiency_summary <- data.frame(
      Metric = "Mutation Efficiency",
      Value = round(eff, 4),
      Treatment_Pattern = treatment_pattern,
      Control_Pattern = control_pattern,
      Treatment_Samples = length(treatment_samples),
      Control_Samples = length(control_samples),
      stringsAsFactors = FALSE
    )
    
    DT::datatable(efficiency_summary, caption = "CRISPR Editing Efficiency")
    
    # Save results
    write.csv(efficiency_summary, 
             file.path(params$outs_path, paste0(treatment_pattern, "_vs_", control_pattern, "_efficiency.csv")),
             row.names = FALSE)
    
  }, error = function(e) {
    message("Error calculating mutation efficiency: ", e$message)
  })
} else {
  message("Cannot calculate mutation efficiency - missing treatment or control samples")
}
```

```{r variant-frequency}
# Calculate and display variant frequencies
if(length(control_samples) > 0) {
  tryCatch({
    variant_freq <- variantFrequency(crispr_set, 
                                   filter.cols = params$control, 
                                   exclude.cols = params$control)
    
    DT::datatable(variant_freq, 
                 caption = paste("Variant frequencies relative to", params$control),
                 options = list(pageLength = 15, scrollX = TRUE))
    
    # Save variant frequencies
    write.csv(variant_freq, 
             file.path(params$outs_path, paste0(params$treatment, "_vs_", params$control, "_variant_frequencies.csv")),
             row.names = TRUE)
    
  }, error = function(e) {
    message("Error calculating variant frequency: ", e$message)
  })
}
```

```{r efficiency-percentages-detailed}
# Get variant counts first
variant_counts <- variantCounts(crispr_set)

# Create the barplot with table to get the result object
bp_result <- barplotAlleleFreqs(variant_counts, include.table = TRUE)

# Percent table matching barplotAlleleFreqs categories from your rownames
vc <- variantCounts(crispr_set)
rn <- rownames(vc)

classify_allele <- function(s){
  s0 <- trimws(s)
  if (identical(s0, "no variant")) return("no variant")
  if (grepl("^SNV:", s0)) return("SNV")
  if (identical(s0, "Other")) return("Other")

  # sum ALL indel lengths from tokens like "...:<len>D" or "...:<len>I"
  m <- gregexpr(":[0-9]+[DI]", s0, perl = TRUE)
  tokens <- regmatches(s0, m)[[1]]
  if (length(tokens) == 0) return("Other")
  L_total <- sum(as.integer(gsub("^:([0-9]+)[DI]$", "\\1", tokens)), na.rm = TRUE)

  if (L_total == 0) return("Other")
  if (L_total %% 3 != 0) {
    if (L_total >= 10) "frameshift indel ≥ 10" else "frameshift indel < 9"
  } else {
    if (L_total > 10) "inframe indel > 10" else "inframe indel ≤ 9"
  }
}

cats <- vapply(rn, classify_allele, character(1))
by_cat <- rowsum(as.matrix(vc), group = cats)
pct_by_cat <- round(sweep(by_cat, 2, colSums(by_cat), "/") * 100, 2)

DT::datatable(as.data.frame(t(pct_by_cat)),
              caption = "Percent of reads by category (per sample)")

# Extract percentage data (try different possible names)
percentage_data <- NULL
for(name in c("percent", "table", "data", "frequencies")) {
  if(name %in% names(bp_result)) {
    percentage_data <- bp_result[[name]]
    message("Found percentage data in component: ", name)
    break
  }
}

if(is.null(percentage_data)) {
  # Fallback: calculate percentages manually
  message("Calculating percentages manually from variant counts")
  total_reads <- colSums(variant_counts)
  percentage_data <- sweep(variant_counts, 2, total_reads, "/") * 100
}

# Create a nice summary table
if(!is.null(percentage_data)) {
  # Add a summary row for each variant type
  if(ncol(percentage_data) > 1) {
    # Calculate mean percentages across samples for each variant
    summary_stats <- data.frame(
      Variant = rownames(percentage_data),
      Mean_Percentage = rowMeans(percentage_data, na.rm = TRUE),
      SD_Percentage = apply(percentage_data, 1, sd, na.rm = TRUE),
      Min_Percentage = apply(percentage_data, 1, min, na.rm = TRUE),
      Max_Percentage = apply(percentage_data, 1, max, na.rm = TRUE),
      stringsAsFactors = FALSE
    )
    
    DT::datatable(summary_stats, 
                 caption = "Variant Frequency Summary Statistics",
                 options = list(pageLength = 10)) %>%
      DT::formatRound(columns = 2:5, digits = 2)
  }
  
  # Display the full percentage table
  DT::datatable(percentage_data, 
               caption = "Allele Frequencies by Sample (%)",
               options = list(pageLength = 15, scrollX = TRUE)) %>%
    DT::formatRound(columns = which(sapply(percentage_data, is.numeric)), digits = 2)
  
  # Save the data
  write.csv(percentage_data, 
           file.path(params$outs_path, paste0(params$treatment, "_vs_", params$control, "_allele_percentages.csv")),
           row.names = TRUE)
  
  if(exists("summary_stats")) {
    write.csv(summary_stats, 
             file.path(params$outs_path, paste0(params$treatment, "_vs_", params$control, "_variant_summary.csv")),
             row.names = FALSE)
  }
}
```

```{r editing-efficiency-summary}
# Create a focused editing efficiency summary
variant_counts <- variantCounts(crispr_set)

# Calculate total reads and editing percentages
editing_summary <- data.frame()

for(sample in colnames(variant_counts)) {
  total_reads <- sum(variant_counts[, sample])
  
  # Assuming first row is "no variant" (reference sequence)
  reference_reads <- variant_counts[1, sample]
  edited_reads <- total_reads - reference_reads
  
  editing_percentage <- (edited_reads / total_reads) * 100
  
  sample_type <- ifelse(grepl(params$treatment, sample), "Treatment", 
                       ifelse(grepl(params$control, sample), "Control", "Other"))
  
  editing_summary <- rbind(editing_summary, data.frame(
    Sample = sample,
    Sample_Type = sample_type,
    Total_Reads = total_reads,
    Reference_Reads = reference_reads,
    Edited_Reads = edited_reads,
    Editing_Percentage = round(editing_percentage, 2),
    stringsAsFactors = FALSE
  ))
}

# Display the editing efficiency summary
DT::datatable(editing_summary, 
             caption = "CRISPR Editing Efficiency Summary",
             options = list(pageLength = 10)) %>%
  DT::formatStyle("Sample_Type",
                 backgroundColor = DT::styleEqual(c("Treatment", "Control"), 
                                                c("lightblue", "lightgreen")))

# Calculate group means
group_summary <- aggregate(Editing_Percentage ~ Sample_Type, editing_summary, 
                          function(x) c(Mean = mean(x), SD = sd(x), N = length(x)))

group_summary_df <- data.frame(
  Group = group_summary$Sample_Type,
  Mean_Editing_Percentage = round(group_summary$Editing_Percentage[, "Mean"], 2),
  SD_Editing_Percentage = round(group_summary$Editing_Percentage[, "SD"], 2),
  Sample_Count = group_summary$Editing_Percentage[, "N"],
  stringsAsFactors = FALSE
)

DT::datatable(group_summary_df, caption = "CRISPR Editing Efficiency by Group")
```


```{r consensus-sequences}
# Display consensus sequences
consensus_seqs <- consensusSeqs(crispr_set)
# Save consensus sequences
write.csv(as.data.frame(consensus_seqs), 
         file.path(params$outs_path, paste0(params$treatment, "_vs_", params$control, "_", params$guide_name, "_consensus_sequences.csv")),
         row.names = TRUE)
```

## Reproducibility

### Code document - `quarto`

The code in quarto document format is included below which can be used to render and rerun this analysis workflow in its entirety. It can be rendered with the `quarto` publishing system running in the R programming environment in the `renv` section. You can download the file using the button below.

```{r qmd-download}
#| echo: false
#| results: "asis"

qmd_embed_download()

```

### R programming environment - `renv`

This project uses `renv` to manage package dependencies. `renv` is an R package that helps create and manage project-specific environments, ensuring that the correct versions of packages are used for reproducibility. To restore the environment, you would typically run `renv::restore()` in your R session, which will install the packages specified in the `renv.lock` file.
You can download the `renv.lock` file below to restore the environment.

```{r renv-download}
#| echo: false
#| results: "asis"

renv_embed_download()

```

The current status of the environment is as follows:

```{r, renv-status}
renv::status()
```

### Session info

```{r session-info}
sessionInfo()
```

---

```{r render-time}
#| echo: false
#| results: "asis"
end_time <- Sys.time()
render_duration <- end_time - start_time

cat("Document render completed at:", format(end_time, "%Y-%m-%d %H:%M:%S"), "\n\n")
cat("Total render time:", round(as.numeric(render_duration, units = "mins"), 2), "minutes\n")
```