---
title: "Analysis: CrispRVariants"
subtitle: "Brzezinski - Purvis Dev paper CRISPR editing outcomes"
author: 
  name: "Michael Kaufman, PhD"
  orcid: 0000-0003-2441-5836
  email: michael.kaufman@cuanschutz.edu
  affiliation: 
    name: "University of Colorado Cancer Center"
    url: "https://medschool.cuanschutz.edu/colorado-cancer-center"
abstract:
  ""

editor: source
theme: cosmo
number-sections: true
link-citations: true
link-external-references: true

format:
  html:
    title-block-banner: "report_assets/analysis-logo.png"
    embed-resources: true
    self_contained: true
    highlight-style: pygments
    code-fold: true
    code-summary: "[code]"
    code-overflow: wrap
    toc: true
    toc-depth: 2
    toc-location: right  
    css: report_assets/custom-toc.css
    grid:
      sidebar-width: 300px
      body-width: 1600px
      margin-width: 300px
      gutter-width: 10px

execute:
  warning: false
  message: false
  fig.align: "center"

params:
  bwa_threads: 6
  outs_path: "outs/01-CrispRVariants"
  fastq_path: "../raw_data/fastq"
  bam_path: "../processed_data"
  ref_genome: "../ref/mm39.fa"
  gtf_file: "../ref/gencode.vM37.basic.annotation.gtf"
  treatment: "TestDHS15primers_R2"
  control: "ControlDHS15primers_R2"
  guide_name: "DHS15g2" # DHS15g1 or DHS15g2
  guide_sequence: "TTTTCTTTTTTATTTAACCG" # GCTGCCCCAGCCTTTCACAA or TTTTCTTTTTTATTTAACCG
  target_chr: "chr14"
  target_padding: 30
---

## Notes

This document is part of the Purvis Dev 2025 CRISPR editing outcomes analysis. It uses the `CrispRVariants` package to analyze CRISPR editing outcomes from paired-end sequencing data. Amplicon sequencing data is used to assess the efficiency of CRISPR editing at a specific target region in the mouse genome.

The following external tools are used in this analysis and must be installed and available in your PATH:
`bwa mem` is used for alignment, and `CrispRVariants` is used to analyze the editing outcomes.
`samtools` is used for BAM file processing.

This experiment uses paired-end sequencing data, where each sample has two FASTQ files (R1 and R2). The analysis will align these reads to a reference genome, create a `CrispRVariants` object, and visualize the editing outcomes.

## Environment

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align='center')

# Start timing the document render
start_time <- Sys.time()
```

### Load libraries

```{r load libraries}

library(CrispRVariants)
library(Rsamtools)
library(GenomicRanges)
library(GenomicFeatures)
library(Biostrings)
library(DT)
library(gridExtra)
library(Gviz)

source("scripts/qmd_embed_download.R")
source("scripts/renv_embed_download.R")

set.seed(87645893)

dir.create(params$outs_path, showWarnings = TRUE)
```

## Analysis parameters

```{r samples}
datatable(t(as.data.frame(params)), 
          options = list(dom = 't', paging = FALSE))
```

## Alignment of FASTQ files

This step calls the BWA and SAMtools commands to align the FASTQ files to the reference genome and create sorted BAM files.

```{r align-fastq}
dir.create(params$bam_path, recursive = TRUE, showWarnings = FALSE)

fastqs <- list.files(params$fastq_path, pattern="\\.(fastq|fq)(\\.gz)?$", full.names=TRUE, ignore.case=TRUE)
stopifnot(length(fastqs) > 0)

for (fq in fastqs) {
  sample <- sub("\\.(fastq|fq)(\\.gz)?$", "", basename(fq), ignore.case=TRUE)
  bam <- file.path(params$bam_path, paste0(sample, "_sorted.bam"))

  if (!file.exists(bam)) {
    message("Aligning: ", sample)
    cmd <- sprintf('bwa mem -t %d "%s" "%s" | samtools sort -@ %d -o "%s"',
                   as.integer(params$bwa_threads), params$ref_genome, fq,
                   as.integer(params$bwa_threads), bam)
    status <- system(cmd)
    if (status != 0) stop("Alignment failed for: ", sample)
    system2("samtools", c("index", "-@", as.character(params$bwa_threads), bam))
  } else {
    message("Skipping ", sample, " — BAM already exists.")
  }
}
```

## CrispR editing outcomes analysis

### Create the target location and reference sequence

```{r auto-detect-target}
# Auto-detect target region based on guide sequence
detect_target_from_guide <- function(ref_genome, guide_seq, target_chr = NULL, flank_size = 50) {
  
  message("Searching for guide sequence: ", guide_seq)
  
  # If chromosome is specified, search only that chromosome
  if(!is.null(target_chr)) {
    search_cmd <- sprintf("samtools faidx %s %s | grep -v '^>' | tr -d '\\n' | grep -o -b '%s\\|%s'",
                         ref_genome, target_chr, guide_seq, 
                         as.character(reverseComplement(DNAString(guide_seq))))
    chromosomes_to_search <- target_chr
  } else {
    # Search all chromosomes - get chromosome list first
    chr_cmd <- sprintf("samtools faidx %s.fai | cut -f1", ref_genome)
    chromosomes_to_search <- system(chr_cmd, intern = TRUE)
    chromosomes_to_search <- chromosomes_to_search[grepl("^chr", chromosomes_to_search)]
  }
  
  target_locations <- list()
  
  for(chr in chromosomes_to_search) {
    message("Searching chromosome: ", chr)
    
    # Extract chromosome sequence
    chr_seq_cmd <- sprintf("samtools faidx %s %s", ref_genome, chr)
    chr_lines <- system(chr_seq_cmd, intern = TRUE)
    
    if(length(chr_lines) < 2) next
    
    chr_seq <- paste(chr_lines[-1], collapse = "")
    chr_dna <- DNAString(chr_seq)
    
    # Search for guide sequence on forward strand
    forward_matches <- matchPattern(guide_seq, chr_dna)
    if(length(forward_matches) > 0) {
      for(i in seq_along(forward_matches)) {
        match_start <- start(forward_matches)[i]
        match_end <- end(forward_matches)[i]
        
        # Calculate cut site (typically 3bp upstream of PAM, which is ~17bp into guide)
        cut_site <- match_start + 17
        
        target_locations[[length(target_locations) + 1]] <- list(
          chr = chr,
          guide_start = match_start,
          guide_end = match_end,
          cut_site = cut_site,
          strand = "+",
          target_start = max(1, cut_site - flank_size),
          target_end = cut_site + flank_size,
          target_loc_forward = flank_size + 1, # Relative to target_start
          target_loc_reverse = flank_size + 1
        )
      }
    }
    
    # Search for guide sequence on reverse strand
    reverse_matches <- matchPattern(reverseComplement(DNAString(guide_seq)), chr_dna)
    if(length(reverse_matches) > 0) {
      for(i in seq_along(reverse_matches)) {
        match_start <- start(reverse_matches)[i]
        match_end <- end(reverse_matches)[i]
        
        # For reverse strand, cut site is typically 3bp downstream
        cut_site <- match_end - 17
        
        target_locations[[length(target_locations) + 1]] <- list(
          chr = chr,
          guide_start = match_start,
          guide_end = match_end,
          cut_site = cut_site,
          strand = "-",
          target_start = max(1, cut_site - flank_size),
          target_end = cut_site + flank_size,
          target_loc_forward = flank_size + 1,
          target_loc_reverse = flank_size + 1
        )
      }
    }
  }
  
  return(target_locations)
}

# Auto-detect target region
detected_targets <- detect_target_from_guide(
  ref_genome = params$ref_genome,
  guide_seq = params$guide_sequence,
  target_chr = params$target_chr,  # Use specified chromosome if available
  flank_size = params$target_padding  # 30bp flanking regions
)

message("Found ", length(detected_targets), " potential target sites")

if(length(detected_targets) == 0) {
  stop("Guide sequence not found in reference genome!")
}

# Display detected targets
detected_df <- do.call(rbind, lapply(detected_targets, function(x) {
  data.frame(
    Chromosome = x$chr,
    Guide_Start = x$guide_start,
    Guide_End = x$guide_end,
    Cut_Site = x$cut_site,
    Strand = x$strand,
    Target_Start = x$target_start,
    Target_End = x$target_end,
    stringsAsFactors = FALSE
  )
}))

DT::datatable(detected_df, caption = "Detected Target Locations for Guide Sequence")

# Use the first detected target (or filter by chromosome if specified)
if(!is.null(params$target_chr)) {
  # Filter to specified chromosome
  chr_matches <- detected_targets[sapply(detected_targets, function(x) x$chr == params$target_chr)]
  if(length(chr_matches) > 0) {
    selected_target <- chr_matches[[1]]
  } else {
    stop("No targets found on specified chromosome: ", params$target_chr)
  }
} else {
  selected_target <- detected_targets[[1]]
}

message("Selected target:")
message("  Chromosome: ", selected_target$chr)
message("  Target region: ", selected_target$target_start, "-", selected_target$target_end)
message("  Strand: ", selected_target$strand)
message("  Cut site: ", selected_target$cut_site)

# Override parameters with detected values
auto_params <- list(
  target_chr = selected_target$chr,
  target_start = selected_target$target_start,
  target_end = selected_target$target_end,
  target_strand = selected_target$strand,
  target_loc_forward = selected_target$target_loc_forward,
  target_loc_reverse = selected_target$target_loc_reverse
)
```

```{r target-location}
# Define the target location and reference sequence as granges
target_location <- GRanges(
  seqnames = params$target_chr,
  ranges = IRanges(start = auto_params$target_start, end = auto_params$target_end),
  strand = auto_params$target_strand
)

# Extract reference sequence
ref_cmd <- sprintf("samtools faidx %s %s:%s-%s",
                  params$ref_genome,
                  seqnames(target_location)[1],
                  start(target_location)[1],
                  end(target_location)[1])

ref_lines <- system(ref_cmd, intern = TRUE)

if(length(ref_lines) < 2) {
  stop("Failed to extract reference sequence")
}

reference_raw <- paste(ref_lines[-1], collapse = "")

# Use reference as-is (don't automatically reverse complement)
reference <- Biostrings::DNAString(reference_raw)

# Validate guide sequence is found in reference
guide_found_forward <- grepl(params$guide_sequence, reference_raw, fixed = TRUE)
guide_found_reverse <- grepl(as.character(Biostrings::reverseComplement(Biostrings::DNAString(params$guide_sequence))), 
                            reference_raw, fixed = TRUE)

if(guide_found_forward) {
  message("Guide sequence found on forward strand in target region")
} else if(guide_found_reverse) {
  message("Guide sequence found on reverse strand in target region")
} else {
  warning("Guide sequence not found in target region - check detection results")
}

message("Target width: ", width(target_location))
message("Reference length: ", nchar(reference_raw))
message("Guide sequence: ", params$guide_sequence)
message("Reference sequence: ", substr(reference_raw, 1, 50), "...")

# Show final parameters being used
final_params_df <- data.frame(
  Parameter = c("target_chr", "target_start", "target_end", "target_strand", 
               "target_loc_forward", "target_loc_reverse"),
  Value = c(auto_params$target_chr, auto_params$target_start, auto_params$target_end, 
           auto_params$target_strand, auto_params$target_loc_forward, auto_params$target_loc_reverse),
  stringsAsFactors = FALSE
)

DT::datatable(final_params_df, caption = "Final Target Parameters")
```

### CrisprSet object

This object is created from the sorted BAM files and the target location. It will contain the read counts and variant information for each sample at the specified target region.

```{r crispr-set}
# Get all sorted BAM files first
all_bam_files <- list.files(params$bam_path, pattern = "_sorted\\.bam$", full.names = TRUE)

if(length(all_bam_files) == 0) {
  stop("No sorted BAM files found in ", params$bam_path, ". Check alignment step.")
}

all_sample_names <- gsub("_sorted\\.bam$", "", basename(all_bam_files))

# Filter to only treatment and control samples based on parameters
treatment_indices <- which(grepl(params$treatment, all_sample_names))
control_indices <- which(grepl(params$control, all_sample_names))

# Combine indices and get corresponding BAM files
selected_indices <- c(treatment_indices, control_indices)
selected_bam_files <- all_bam_files[selected_indices]
selected_sample_names <- all_sample_names[selected_indices]

message("All BAM files found: ", length(all_bam_files))
message("Treatment pattern '", params$treatment, "' matches: ", length(treatment_indices), " files")
message("Control pattern '", params$control, "' matches: ", length(control_indices), " files")
message("Selected BAM files for analysis: ", length(selected_bam_files))

if(length(selected_bam_files) == 0) {
  stop("No BAM files match the treatment (", params$treatment, ") or control (", params$control, ") patterns!")
}

message("Selected BAM files:")
for(i in seq_along(selected_bam_files)) {
  sample_type <- if(i <= length(treatment_indices)) "TREATMENT" else "CONTROL"
  message("  - ", basename(selected_bam_files[i]), " (", sample_type, ")")
}

# Check selected files exist and are indexed
for(bam_file in selected_bam_files) {
  if(!file.exists(bam_file)) {
    stop("BAM file not found: ", bam_file)
  }
  if(!file.exists(paste0(bam_file, ".bai"))) {
    message("Indexing missing for: ", basename(bam_file))
    indexBam(bam_file)
  }
}

# Use ONLY the selected BAM files for CrisprSet creation
crispr_set <- readsToTarget(selected_bam_files, 
                           target = target_location, 
                           reference = reference,
                           names = selected_sample_names, 
                           target.loc = auto_params$target_loc_forward)

DT::datatable(variantCounts(crispr_set))

# Update the global variables to use selected samples
sorted_bam_files <- selected_bam_files
bam_sample_names <- selected_sample_names
```

### Variant plots

This section creates variant plots for the CRISPR editing outcomes. It uses the `CrispRVariants` package to visualize the variants detected in the target region across samples.

```{r variant-plots}
gtf_fname <- params$gtf_file

# Check if GTF file exists
if(file.exists(gtf_fname)) {
  message("Creating TxDb from GTF file...")
  txdb <- GenomicFeatures::makeTxDbFromGFF(gtf_fname, format = "gtf")
  
  # Check if there are any genes in our target region
  target_genes <- genes(txdb, filter = list(tx_chrom = seqnames(target_location)[1]))
  overlapping_genes <- subsetByOverlaps(target_genes, target_location)
  
  if(length(overlapping_genes) == 0) {
    message("No genes found overlapping target region. Proceeding without gene annotation.")
    txdb <- NULL
  } else {
    message("Found ", length(overlapping_genes), " genes overlapping target region")
    print(overlapping_genes)
  }
} else {
  message("GTF file not found at: ", gtf_fname)
  message("Proceeding without gene annotation")
  txdb <- NULL
}
```

```{r variant-plots2}
#| fig.width: 16
#| fig.height: 16
#| results: "hide"
#| message: false
#| warning: false

# Now group will only include the selected samples
group <- ifelse(grepl(params$treatment, bam_sample_names), "treatment", 
                ifelse(grepl(params$control, bam_sample_names), "control", "other"))

# Validation - should now only show selected samples
message("Sample groupings (selected samples only):")
message("Treatment (", params$treatment, "): ", paste(bam_sample_names[group == "treatment"], collapse = ", "))
message("Control (", params$control, "): ", paste(bam_sample_names[group == "control"], collapse = ", "))
message("Other: ", paste(bam_sample_names[group == "other"], collapse = ", "))

# Verify we have both groups
if(!any(group == "treatment")) {
  warning("No treatment samples found in selected BAMs!")
}
if(!any(group == "control")) {
  warning("No control samples found in selected BAMs!")
}

plotVariants(crispr_set, txdb = txdb, gene.text.size = 8, main = paste0(params$treatment, " vs ", params$control),
    row.ht.ratio = c(1,8), col.wdth.ratio = c(4,2),
    plotAlignments.args = list(line.weight = 0.5, ins.size = 2, 
                               legend.symbol.size = 4),
    plotFreqHeatmap.args = list(plot.text.size = 3, x.size = 8, group = group, 
                                legend.text.size = 8, 
                                legend.key.height = grid::unit(0.5, "lines"))) 

# Use parameter-based filename
pdf(file.path(params$outs_path, paste0(params$treatment, "_vs_", params$control, "_variant_plot.pdf")), 
    width = 16, height = 16)
plotVariants(crispr_set, main = paste0(params$treatment, " vs ", params$control))
dev.off()
```

```{r mutation-efficiency}
# Calculate mutation efficiency
control_pattern <- params$control
treatment_pattern <- params$treatment

# Dynamic sample identification using the parameters
treatment_samples <- bam_sample_names[grepl(treatment_pattern, bam_sample_names)]
control_samples <- bam_sample_names[grepl(control_pattern, bam_sample_names)]

message("Treatment samples: ", paste(treatment_samples, collapse = ", "))
message("Control samples: ", paste(control_samples, collapse = ", "))

if(length(treatment_samples) > 0 && length(control_samples) > 0) {
  tryCatch({
    eff <- mutationEfficiency(crispr_set, 
                             filter.cols = control_pattern, 
                             exclude.cols = control_pattern)
    message("Mutation efficiency: ", round(eff, 4))
    
    # Create summary table
    efficiency_summary <- data.frame(
      Metric = "Mutation Efficiency",
      Value = round(eff, 4),
      Treatment_Pattern = treatment_pattern,
      Control_Pattern = control_pattern,
      Treatment_Samples = length(treatment_samples),
      Control_Samples = length(control_samples),
      stringsAsFactors = FALSE
    )
    
    DT::datatable(efficiency_summary, caption = "CRISPR Editing Efficiency")
    
    # Save results
    write.csv(efficiency_summary, 
             file.path(params$outs_path, paste0(treatment_pattern, "_vs_", control_pattern, "_efficiency.csv")),
             row.names = FALSE)
    
  }, error = function(e) {
    message("Error calculating mutation efficiency: ", e$message)
  })
} else {
  message("Cannot calculate mutation efficiency - missing treatment or control samples")
}
```

```{r variant-frequency}
# Calculate and display variant frequencies
if(length(control_samples) > 0) {
  tryCatch({
    variant_freq <- variantFrequency(crispr_set, 
                                   filter.cols = params$control, 
                                   exclude.cols = params$control)
    
    DT::datatable(variant_freq, 
                 caption = paste("Variant frequencies relative to", params$control),
                 options = list(pageLength = 15, scrollX = TRUE))
    
    # Save variant frequencies
    write.csv(variant_freq, 
             file.path(params$outs_path, paste0(params$treatment, "_vs_", params$control, "_variant_frequencies.csv")),
             row.names = TRUE)
    
  }, error = function(e) {
    message("Error calculating variant frequency: ", e$message)
  })
}
```

### Consensus sequences

```{r consensus-sequences}
# Display consensus sequences
consensus_seqs <- consensusSeqs(crispr_set)
DT::datatable(as.data.frame(consensus_seqs), 
             caption = "Consensus Sequences by Sample",
             options = list(scrollX = TRUE))

# Save consensus sequences
write.csv(as.data.frame(consensus_seqs), 
         file.path(params$outs_path, paste0(params$treatment, "_vs_", params$control, "_consensus_sequences.csv")),
         row.names = TRUE)
```

## Reproducibility

### Code document - `quarto`

The code in quarto document format is included below which can be used to render and rerun this analysis workflow in its entirety. It can be rendered with the `quarto` publishing system running in the R programming environment in the `renv` section. You can download the file using the button below.

```{r qmd-download}
#| echo: false
#| results: "asis"

qmd_embed_download()

```

### R programming environment - `renv`

This project uses `renv` to manage package dependencies. `renv` is an R package that helps create and manage project-specific environments, ensuring that the correct versions of packages are used for reproducibility. To restore the environment, you would typically run `renv::restore()` in your R session, which will install the packages specified in the `renv.lock` file.
You can download the `renv.lock` file below to restore the environment.

```{r renv-download}
#| echo: false
#| results: "asis"

renv_embed_download()

```

The current status of the environment is as follows:

```{r, renv-status}
renv::status()
```

### Session info

```{r session-info}
sessionInfo()
```

---

```{r render-time}
#| echo: false
#| results: "asis"
end_time <- Sys.time()
render_duration <- end_time - start_time

cat("Document render completed at:", format(end_time, "%Y-%m-%d %H:%M:%S"), "\n\n")
cat("Total render time:", round(as.numeric(render_duration, units = "mins"), 2), "minutes\n")
```